---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    keep_md: yes
    self_contained: no
  pdf_document: default
---
## Set up the environment -- load libraries required
```{r}
library("dplyr")
library("lattice")
```

## Loading and preprocessing the data

Read into data table:

```{r}
fit_table<-read.csv(unz("activity.zip","activity.csv"))
```

## What is mean total number of steps taken per day?
Strategy: use aggregate() to compute a data frame of sums by day, then use this data for the histogram plot and average/median computations.

```{r}
## total_steps contains the total number of steps taken each day 
##    (summed across all 5 min time intervals)

## sum up all observations for each day, ignoring NA values
total_steps <- aggregate(fit_table$steps,by=list(fit_table$date),sum)


## create a histogram showing the frequency distribution of 
##    daily step totals
hist(total_steps$x, breaks=20, main="Histogram of Average Steps per Day",
     xlab="Daily Step Total")

## compute and display the arithmetic average of daily step totals
cat("Average Daily Step Total: ", mean(total_steps$x,na.rm=TRUE))

## compute and display the median daily step total
cat("Median Daily Step Total: ", median(total_steps$x,na.rm=TRUE))
```

## What is the average daily activity pattern?
Strategy: Use aggregate() to compute a data frame of means by interval, then plot this data.

```{r}
## steps_by_interval contains the average number of steps in each time 
##    interval across the 61 days in the data

## average steps in a given time interval across all days in the data set, 
##    ignoring NA values
steps_by_interval <- aggregate(fit_table$steps,by=list(fit_table$interval), 
                               mean, na.rm=TRUE)

plot(steps_by_interval$Group.1, steps_by_interval$x,type="l", xlab="Interval",
     ylab="Avg Steps per Interval", main="Daily Activity Pattern")
## compute maximum average steps by interval and output both number and interval
max_int <- max(steps_by_interval$x)
max_time <- steps_by_interval[steps_by_interval$x==max_int,]
cat("maximum steps per interval is ", max_int, 
    " recorded in interval ", as.integer(max_time[1,1]))
```

## Imputing missing values
Strategy: Use the average steps in each interval as the imputed value for each NA in the same interval.

```{r}
## calculate number of missing values

missing_data <- is.na(fit_table$steps)
cat("number of missing values", sum(missing_data))

## use dplyr to manipulate data when imputing missing values

## create new table that contains original data plus interval means
int1<- tbl_df(left_join(fit_table,
           as_data_frame(list(interval=steps_by_interval$Group.1, 
           mean_steps=steps_by_interval$x))))

## add variable with imputed steps
## step one -- replace NA with zero
steps_z<-fit_table$steps
steps_z[is.na(steps_z)]<-0
## step 2 add imputed steps as steps_i variable to table
int2<-mutate(int1, steps_i=steps_z+mean_steps*missing_data)

## create new data frame identical to original except for imputed NAs
new_fit_table <- data.frame(steps=int2$steps_i,date=int2$date,
                           interval=int2$interval)

##compute new total steps
total_steps2 <- aggregate(new_fit_table$steps,by=list(new_fit_table$date),sum)

## create a histogram showing the frequency distribution of 
##    daily step totals
hist(total_steps2$x, breaks=20, main="Histogram of Average Steps per Day",
     xlab="Daily Step Total")

## compute and display the arithmetic average of daily step totals
cat("Average Daily Step Total: ", mean(total_steps2$x))

## compute and display the median daily step total
cat("Median Daily Step Total: ", median(total_steps2$x))

```

NAs were replaced with mean for that time interval over the entire data set. The NAs were all clustered such that there were 8 days with only NAs and no NAs outside of those 8 days.  The effect of this strategy of filling in NAs was to remove 8 days of zero steps and add 8 days with a total step count equal to the sum of the means by interval (which, by construction is the same as the mean of the non-missing days). Thus there was no change to the mean.  Since the 8 days were a large proportion of the data (8/61) and were clustered in the middle of the data, the sum of the interval means became the median daily total.

## Are there differences in activity patterns between weekdays and weekends?
Strategy: create a function to return "Weekday" and "Weekend" as appropriate

Lattice library used for panel plot

```{r}
## function to be called in next step
wkend<-function(dotw){
        if(dotw=="Sat" | dotw=="Sun") wkend<-"Weekend" else wkend<-"Weekday"
         wkend
}

## add a column to the data frame with the factor Weekday/Weekend
int1<-mutate(new_fit_table,wkfac=sapply(weekdays(strptime(new_fit_table$date,
                  format="%Y-%m-%d"), abbreviate=TRUE),wkend))

## average steps in a given time interval across all days in the data set, 
##      separately for weekdays and weekends 
steps_by_interval2 <- aggregate(int1$steps,by=list(int1$wkfac, int1$interval), 
                                mean)
## use attach to so I can refer to just the elements of steps_by_interval2
attach(steps_by_interval2)
## panel plot using Lattice
xyplot(x~Group.2|Group.1,type="l",xlab="Interval",
     ylab="Avg Steps per Interval", main="Daily Activity Pattern", 
     layout=c(1,2),scales=list(y=list(alternating=FALSE)))
```

